Can Natural Models Simplify The Metatheory of Type Theory in Cubical Agda?

### 2025-05-20 tex/FSCD25/rebuttal.txt

We thank the reviewers for their detailed and constructive feedback.

Reviewer 1:

>Lack of focus on metaproperties about object type theory

The interpretation into the standard model is described in Section 4.2, albeit
very briefly due to a lack of space. But it demonstrates that our approach can
remove noise in definitions and proofs, with the details available in the Agda
formalisation, as stated in Section 4.2.

>Practical effectiveness of the approach

We didn't manage to remove all transports in the case of inductive types, as we
acknowledge in Section 3.5. However, we still improve on the number of
transports needed for types and introduction forms.  We concentrate on
eliminating transports along type substitutions, which introduce overhead early
in the formalisation. Addressing this specific source of transports hopefully
serves as a use case of QIIRTs and a step towards a solution to the transport
hell. Our preliminary substitution calculus NbE experiment that the reviewer
mentions indicate that the QIIRT-based definition reduces the code size by
approximately 20% (from 229 to 182 LoC), for this simple case.

>It seems like [your GitHub repository] contains more, such as work-in-progress
>proof of NbE. Is there a reason to exclude that from your future work?

The current experiments are small in terms of code size, and thus we didn't
think it was worthwhile to mention them in the paper. But we of course hope to
turn them into larger projects at a later stage. Since the QIIRT definition is
(syntactically) isomorphic to the QIIT one, and the eliminator only introduces
additional derived equations, we expect existing meta-theoretical results based
on QIITs to transfer with minimal changes.

Reviewer 2:

>Do you have any (even vague) idea of how your generalisation of IRT would fare
>with any of the justification for induction-recursion?

It seems plausible that Dybjer and Setzer's set-theoretic model ([20] in our
paper) could be adapted to justify this generalisation. If so, it should also
be straightforward to extend this model to also include equality constructors,
since quotients are easy in set theory.

>Can you explain how you derive your notion of eliminator?
>In particular, what (if any) is the difference between that eliminator and
>that for the standard QIIT?
>And do you have a justification/intuition for why you require that your model
>satisfies only propositionally the rules which are definitional in the QIIRT?

The eliminator for a QIIRT is derived following the general QIIT schema, where
each function clause in the recursive part of the QIIRT definition corresponds
to an equality constructor.

Since we implement some constructors as recursive functions rather than
constructors, the associated computation rules become propositional equalities
in the eliminator (Proposition 4.2). These propositional equalities cannot be
stated as definitional equalities, since there is no way to talk about
definitional equality within type theory.

For the same reason, models cannot be required to satisfy definitional equalities for
these clauses.

>Can you explain why you are not able to have a notion of term-level
>substitution defined solely as a recursive function?
>Why you chose to have composition and other operation not be themselves
>defined? And why substitution is defined by induction on the structure of
>substitution, rather than on that of terms?

We do not define term-level substitution recursively, as doing so does not seem
to reduce the number of transports needed, beyond the "smart constructor"
[–]Tm– (Section 3.3). Defining more operations recursively would increase the
number of proof obligations such as verifying confluence.

The smart constructor is defined by induction on the structure of substitution,
because the clauses based on the substitution structure are exactly those needed
to align with the definitional behaviour of type substitution.

>Can you comment on the relation with research on explicit substitution
>calculi? In particular, do you view any fundamental obstruction to/reason not
>to adapt the complete rewriting system of Schäfer, Smolka and Tebbi, which
>seems to solve your issue in section 3.5, to a typed setting?

The work of Schäfer et al. employs a normalisation function into normal
expressions (see Section 6 and Theorem 32, op. cit.). In our setting, this
would correspond to replacing equality constructors with a normalisation
function, similar to the QIRT definition of join lists in Section 2.2 and the
work by Bense et al. Adapting their approach to a type theory with universes
and inductive types seems useful but likely requires a definition that departs
far from the typical CwF-based definition of type theory.

Reviewer 3:

>In your formalisation there is less transport hell, but is there any kind of
>new hell? E.g. "implicit argument hell" [...].

Indeed, some implicit arguments must be given explicitly in the QIIRT version.
Compared to the work saved from dealing with transports explicitly, we believe
it is a good deal, since inserting implicit arguments does not complicate the
equational theory.

### 2025-09-02 tex/CPP26/main.tex
[ tex ] outline

%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigplan')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigplan,10pt,review]{acmart}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}
%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2018/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.

\usepackage{xspace}
\newcommand{\CA}{\textsc{Cubical Agda}\xspace}
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Formalising type theory in a cubical language}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
This paper presents a formalisation of a core type theory and its metatheory within \CA.
The type theory features $\Pi$-types, an inductive type of Booleans, and a universe.
We formalise key meta-properties, including standard and logical predicate interpretations, canonicity, and normalisation by evaluation for typed terms.

The type theory is defined intrinsically as a representable map of presheaves using quotient inductive-inductive-recursive types.
Moreover, type theory and its elimination are defined with dependent identities without transports.
This work is the first formalization of type theory that uses \CA's built-in support for quotient types, without using Licata's workaround and any rewriting rules, whereas its proofs of canonicity and normalisation are computational.
\end{abstract}

@@

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
%\bibliography{sample-base}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
lacinia dolor. Integer ultricies commodo sem nec semper.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
eros. Vivamus non purus placerat, scelerisque diam eu, cursus
ante. Etiam aliquam tortor auctor efficitur mattis.

\end{document}
\endinput
%%
%% End of file `sample-sigplan.tex'.

>>>>

\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXXXXX.XXXXXXX}
%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2018/06}

%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{xspace}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Macros %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\CA}{\textsc{Cubical Agda}\xspace}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Can Natural Models Simplify The Metatheory of Type Theory in Cubical Agda?}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}

  We formalise type theory in type theory, using Awodey's notion of \emph{natural model} of type theory.
  The initial natural model can be represented using quotient inductive-inductive-recursive types in the proof assistant \CA.
Using natural models leads to fewer instances of so-called ``transport hell'' in the syntax of type theory and its elimination rules.
To show that the approach is feasible, we formalise some meta-properties, including the standard % and logical predicate
interpretation, normalisation by evaluation for typed terms, and strictification constructions.
Since our formalisation is carried out using \CA's built-in support for quotient types, all our constructions compute.
However, the use of transports along equations often reappeared when developing more sophisticated metatheory.
Hence it is still a considerable struggle to develop the metatheory of type theory using welltyped terms with current proof assistant technology, and the effort is about the same using natural models or not.

%   This paper presents a formalisation of a core type theory and its metatheory within \CA.
% The type theory features $\Pi$-types, an inductive type of Booleans, and a universe.
% We formalise key meta-properties, including standard and logical predicate interpretations, canonicity, and normalisation by evaluation for typed terms.

% The type theory is defined intrinsically as a representable map of presheaves using quotient inductive-inductive-recursive types.
% Moreover, type theory and its elimination are defined with dependent identities without transports.
% This work is the first formalization of type theory that uses \CA's built-in support for quotient types, without using Licata's workaround and any rewriting rules, whereas its proofs of canonicity and normalisation are computational.
\end{abstract}

@@

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle
\bibliographystyle{ACM-Reference-Format}

\section{Introduction}
% FNF (Fri 5 Sep)

Transport hell = transports appear inside terms that you actually want to study, hence you need to do a lot of reasoning about transport rather than the actual objects of interest.

\section{Setting and metatheory}
% FNF (Sun 7 Sep)

Cubical Agda with UIP

QIIRTs

\section{Type theory as a natural model}
% LTC (Sun 7 Sep)

\subsection{Natural models}

\subsection{The Syntax of Type Theory as a Quotient Inductive-Inductive-Recursive Type}

QIIRT def of the syntax

\subsection{Recursion and Elimination Principles}

\subsection{Strictification}

\section{Metatheory}
% LTC (Tue 9 Sep)

\subsection{Standard Model}

% Argue for relative consistency?

\subsection{Normalisation by Evaluation}

% (for SC)

\section{Comparison with Other Approaches}
% FNF (Tue 9 Sep)

Compared to QIIT:

\begin{itemize}
\item Fewer transports in the syntax, but they tend to come back in concrete models
\item Strictification orthogonal
\end{itemize}

Compared to untyped version:
\begin{itemize}
\item Untyped version might still be easiest to work with, with current proof assistant technology
\end{itemize}

\section{Discussion}
% LTC (at first)

Practical considerations (eg NBE computes, support for interleaved mutual definitions)

Support from proof assistant (eg safety guarantees)

Easier in OTT-based proof assistant

\end{document}

%\begin{acks}
%To Robert, for the bagels and explaining CMYK and color spaces.
%\end{acks}
